On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-12-05 paper.tex
Rewriting

The expression \eqref{eq: theo: funl, form; funl 2} allows a neat interpretation of $\funl(\tiles)$ (as stems from the arguments used in the proof of Theorem~\ref{theo: funl, form}). Namely, $\funl^2(\tiles)$ is the sum of the two terms that appear in that expression. The first term is the squared distance from $(2a, 2b)$ to the diagonal line defined by $i+j-1=t$; and the second term is additional squared distance incurred from rounding $i$, $j$ to integer values.

@@

From Theorem~\ref{theo: funt, sq, form} it stems that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. In addition, noting that $\funt(\len) = \tiles$ is equivalent to $\funl(\tiles) < \len \leq \funl(\tiles+1)$ and using Corollary~\ref{theo: funl, sq, form} the following characterization of $\funt$ is obtained. For $\tiles \geq 3$ with $\tiles$ odd, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-1)^2+(\tiles-3)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $\tiles$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\len \in \left(\frac{ \sqrt{(\tiles-2)^2+(\tiles-4)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}
% *** The corollary says \sqrt{\frac{(\tiles-3)^2+1} {2}}, which comes from the corresponding theorem. This is equal to \frac{\sqrt{(\tiles-2)^2+(\tiles-4)^2}} {2}. Here we are using the second expression, which is probably needed for the random analysis. If so, the expression in the corollary should be changed from the first to the second. (But this change cannnot be made in the theorem it comes from. It's hopefully not a problem that the expression changes from the theorem to the corollary).

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\tiles = \funt(\len)$ is
\begin{equation}
\label{eq: probmax} %***Change order of the two summands under each square root here and in the rest of the proof, in accordance with \eqref{eq: square, odd max tiles, lengths}. (Or change there, but I think that is less preferable because it would correspond to using the y coordinate, then the x coordinate from the figure)
\begin{split}
\probmax(\len) & = g\left(\len, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\len, \tiles-3, \tiles-3 \right) + 2 g\left(\len, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \len \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) + 2 g\left(\len, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given $\tiles = \funt(\len)$, with $\tiles$ odd. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

@@

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles that can contain the second endpoint of a segment that visits the maximum number of tiles $\tiles= \funt(\len)$, with $\tiles$ odd. Example for $\tiles = 5$%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

@@

The technique used in the proof of Theorem~\ref{theo: probmax, sq} could in principle be used for obtaining the probability that the number of tiles visited by the segment equals or exceeds any given value $t < \funt(\len)$. The resulting expressions, however, are cumbersome, because more cases (more subintervals for the length) need to be distinguished depending on how many tiles with lower-left corner $(i-1,j-1)$ such that $i+j-1=t$ can be reached from the reference tile.

\begin{figure}%
\centering%
\includegraphics[width=.75\textwidth]{probmax_len}%
\caption{Probability $\probmax(\len)$ that a random segment of length $\len$ visits the maximum number of tiles%
}%
\label{fig: probmax_len}%
\end{figure}%

Figure~\ref{fig: probmax_len} shows the probability $\probmax(\len)$ of visiting the maximum number of tiles on a unit square grid, as given by \eqref{eq: probmax}. As $\len$ grows, $\probmax(\len)$ has a jump when the maximum number of tiles that can be visited increases by $1$ (and there the function takes the value approached from the left). This happens when $\len$ equals the right endpoint of the interval \eqref{eq: square, odd max tiles, lengths} for $\tiles$ odd or of \eqref{eq: square, even max tiles, lengths} for $\tiles$ even. Let these length values be denoted as $\len_\tiles$:
\begin{equation}
\len_\tiles = \begin{cases}
\displaystyle
\frac{\sqrt{(\tiles-3)^2 + (\tiles-1)^2}} {2} & \text{ for $\tiles$ odd, $\tiles \geq 3$} \\
\displaystyle
\frac{\tiles-2}{\sqrt{2}} & \text{ for $\tiles$ odd, $\tiles \geq 4$}.
\end{cases}
\end{equation}
For example, the first continuous section seen in Figure~\ref{fig: probmax_len} corresponds to a maximum number of tiles $\tiles=3$, for lengths in the interval $(0, \len_3]$, where $\len_3= 1$. The second corresponds to $\tiles=4$, for lengths in $(\len_3, \len_4]$, where $\len_4= \sqrt{2}$. Within each continuous section the probability monotonically increases from $0$ to a maximum value. The heights of the maxima follow different trends for odd and even $\tiles$ respectively, as can be seen in the figure; but in each case they are asymptotically proportional to $1/\len_\tiles$. This is established by the next result.

\begin{proposition}
The function $\probmax$ has the following asymptotic behaviour:
\begin{align}
\label{eq: probmax liminf}
\liminf_{\len \rightarrow \infty}\, \probmax(\len) & = 0 \\ 
\label{eq: probmax lim odd}
\lim_{\substack{\tiles \rightarrow \infty \\
\tiles \ \mathrm{odd}}}\, \tiles \probmax(\len_\tiles) & = \frac 2 \pi \\
\label{eq: probmax lim even}
\lim_{\substack{\tiles \rightarrow \infty \\ \tiles \ \mathrm{even}}}\, \tiles \probmax(\len_\tiles) & = \frac 8 {3\pi} \\
\label{eq: probmax limsup}
\limsup_{\len \rightarrow \infty}\, \len \probmax(\len) & = \frac {4\sqrt{2}}{3\pi}.
\end{align}
\end{proposition}

>>>>

The expression \eqref{eq: theo: funl, form; funl 2} allows a neat interpretation of $\funl(\tiles)$ (as stems from the arguments used in the proof of Theorem~\ref{theo: funl, form}). Namely, $\funl^2(\tiles)$ is the sum of the two terms that appear in that expression. The first term is the squared distance from $(2a, 2b)$ to the diagonal line defined by $i+j-1 = \tiles$; and the second term is additional squared distance incurred from rounding $i$, $j$ to integer values.

@@

From Theorem~\ref{theo: funt, sq, form} it stems that odd values of $\funt(\len)$ correspond to $\isoli = \jsoli$, whereas even values are achieved with $\jsoli = \isoli+1$. In addition, noting that $\funt(\len) = \tiles$ is equivalent to $\funl(\tiles) < \len \leq \funl(\tiles+1)$ and using Corollary~\ref{theo: funl, sq, form} the following characterization of $\funt$ is obtained. For $\tiles \geq 3$ with $\tiles$ odd, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, odd max tiles, lengths}
\len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{ \sqrt{(\tiles-1)^2+(\tiles-3)^2}}{2} \right].
\end{equation}
For $\tiles \geq 4$, $\tiles$ even, $\funt(\len) = \tiles$ if and only if
\begin{equation}
\label{eq: square, even max tiles, lengths}
\len \in \left(\frac{ \sqrt{(\tiles-2)^2+(\tiles-4)^2}}{2}, \frac{\tiles-2}{\sqrt{2}} \right].
\end{equation}
% *** The corollary says \sqrt{\frac{(\tiles-3)^2+1} {2}}, which comes from the corresponding theorem. This is equal to \frac{\sqrt{(\tiles-2)^2+(\tiles-4)^2}} {2}. Here we are using the latter expression, which is probably needed for the random analysis. If so, the expression in the corollary should be changed from the first to the second. (But this change cannnot be made in the theorem it comes from. It's hopefully not a problem that the expression changes from the theorem to the corollary).

@@

\begin{theorem}
\label{theo: probmax, sq}
Given $\len \in \mathbb R^+$, consider a unit square grid and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\tiles = \funt(\len)$ is
\begin{equation}
\label{eq: probmax} %***Change order of the two summands under each square root here and in the rest of the proof, in accordance with \eqref{eq: square, odd max tiles, lengths}. (Or change there, but I think that is less preferable because it would correspond to using the y coordinate, then the x coordinate from the figure)
\begin{split}
\probmax(\len) & = g\left(\len, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } \tiles \text{ odd, } \textstyle \len \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right] \\
\probmax(\len) & = g\left(\len, \tiles-3, \tiles-3 \right) + 2 g\left(\len, \tiles-5, \tiles-1 \right) \\
& \text{for } \tiles \text{ odd, } \textstyle \len \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right] \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) \quad \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right] \\
\probmax(\len) & = 2 g\left(\len, \tiles-4, \tiles-2 \right) + 2 g\left(\len, \tiles-6, \tiles \right) \\
& \text{for } \tiles \text{ even, } \textstyle \len \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

@@

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-1)/2,(\tiles-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((\tiles-3)/2,(\tiles+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions for a given $\tiles = \funt(\len)$ on a unit square grid, with $\tiles$ odd. Example for $\tiles = 7$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

@@

\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Possible tiles for the endpoints of a segment that visits the maximum number of tiles $\tiles= \funt(\len)$ on a unit square grid, with $\tiles$ odd. Example for $\tiles = 5$%
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

@@

The technique used in the proof of Theorem~\ref{theo: probmax, sq} could in principle be used for obtaining the probability that the number of tiles visited by the segment equals or exceeds any given value $t < \funt(\len)$. The resulting expressions, however, are cumbersome, because more cases (more subintervals for the length) need to be distinguished depending on how many tiles with lower-left corner $(i-1,j-1)$ such that $i+j-1=\tiles$ can be reached from the reference tile.

\begin{figure}%
\centering%
\includegraphics[width=.75\textwidth]{probmax_len}%
\caption{Probability $\probmax(\len)$ that a random segment of length $\len$ visits the maximum number of tiles on a unit square grid%
}%
\label{fig: probmax_len}%
\end{figure}%

Figure~\ref{fig: probmax_len} shows the probability $\probmax(\len)$ of visiting the maximum number of tiles on a unit square grid, as given by \eqref{eq: probmax}. As $\len$ grows, $\probmax(\len)$ has a jump when $\funt(\len)$ increases by $1$ (and there $\probmax(\len)$ takes the value approached from the left). This happens when $\len$ equals the right endpoint of the interval \eqref{eq: square, odd max tiles, lengths} for $\tiles$ odd or of the interval \eqref{eq: square, even max tiles, lengths} for $\tiles$ even. These length values will be denoted as $\len_\tiles$:
\begin{equation}
\len_\tiles = \begin{cases}
\displaystyle
\frac{\sqrt{(\tiles-3)^2 + (\tiles-1)^2}} {2} & \text{ for $\tiles$ odd, $\tiles \geq 3$} \\
\displaystyle
\frac{\tiles-2}{\sqrt{2}} & \text{ for $\tiles$ odd, $\tiles \geq 4$},
\end{cases}
\end{equation}
and satisfy $\funt(\len_\tiles) = \tiles$.

The first continuous section seen in Figure~\ref{fig: probmax_len} corresponds to a maximum number of visited tiles $\tiles=3$, for lengths in the interval $(0, \len_3]$, where $\len_3= 1$. The second corresponds to $\tiles=4$, for lengths in $(\len_3, \len_4]$, where $\len_4= \sqrt{2}$. Within each continuous section the probability monotonically increases from $0$ to a maximum value. The heights of the maxima follow different trends for odd and even $\tiles$ respectively, as can be seen in the figure; but in each case they are asymptotically proportional to $1/\len_\tiles$. This is established by the next result.

\begin{proposition}
The function $\probmax$ has the following asymptotic behaviour:
\begin{align}
\label{eq: probmax liminf}
\liminf_{\len \rightarrow \infty}\, \probmax(\len) & = 0, \\ 
\label{eq: probmax lim odd}
\lim_{\substack{\tiles \rightarrow \infty \\
\tiles \ \mathrm{odd}}}\, \tiles \probmax(\len_\tiles) & = \frac 2 \pi, \\
\label{eq: probmax lim even}
\lim_{\substack{\tiles \rightarrow \infty \\ \tiles \ \mathrm{even}}}\, \tiles \probmax(\len_\tiles) & = \frac 8 {3\pi}, \\
\label{eq: probmax limsup}
\limsup_{\len \rightarrow \infty}\, \len \probmax(\len) & = \frac {4\sqrt{2}}{3\pi}.
\end{align}
\end{proposition}
